---
title: "Practice MaxLike Method"
author: "Natasha Brown"
date: "November 28th, 2018"
output: 
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    fig_width: 8
    fig_heigth: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Carolina Wren data used by Royle et al. (2012)

These data come from the North American Breeding Bird Survey.

[*Royle, J.A., R.B. Chandler, C. Yackulic and J. D. Nichols. 2012. Likelihood analysis of species occurrence probability from presence-only data for modelling species distributions. Methods in Ecology and Evolution. doi: 10.1111/j.2041-210X.2011.00182.x*](Royle_et_al-2012-Methods_in_Ecology_and_Evolution.pdf) 


```{r carw, message=FALSE,echo=FALSE}
library(maxlike)
data(carw)
# Convert data frame to a list of rasters 
rl <- lapply(carw.data$raster.data, function(x) { 
  m <- matrix(x, nrow=carw.data$dim[1], ncol=carw.data$dim[2], byrow=TRUE) 
  r <- raster(m)
  extent(r) <- carw.data$ext 
  r
})
# Stack and name them 
rs <- stack(rl[[1]], rl[[2]], rl[[3]], rl[[4]], rl[[5]], rl[[6]]) 
names(rs) <- names(carw.data$raster.data)
plot(rs)
```

## Fake data for the Maunga Whau volcano

Based upon volcano in the datasets package with no coordinate system attributes

```{r Maungawhau, message=FALSE,echo=FALSE}
data(MaungaWhau) 
elev <- raster(MaungaWhau$elev, xmn=0, xmx=61, ymn=0, ymx=87) 
precip <- raster(MaungaWhau$precip, xmn=0, xmx=61, ymn=0, ymx=87)
plot(elev)
plot(precip)
```

## Model occurrence probability using presence-only data

"This function estimates the probability of occurrence using presence-only data and spatially-referenced covariates. Species distribution maps can be created by plotting the expected values of occurrence probability. The model is described by [Royle et al. (2012)](https://besjournals.onlinelibrary.wiley.com/doi/epdf/10.1111/j.2041-210X.2011.00182.x)."


**Warning:** "Maximizing the log-likelihood function is achieved using the optim function, which can fail to ﬁnd the global optima if sensible starting values are not supplied. The default starting values are rep(0, npars), which will often be adequate if the covariates have been standardized. Standardizing covariates is thus recommended. Even when covariates are standardized, it is always a good idea to try various starting values to see if the log-likelihood can be increased. When ﬁtting models with many parameters, good starting values can be found by ﬁtting simpler models ﬁrst".

**Note: 1.)**"Points and rasters should the same coordinate system. The program does not check this so it is up to the user.**2.)**The MAXENT software assumes that species prevalence is known a priori. If the user does not specify a value for prevalence, prevalence is set to 0.5. MAXENT predictions of occurrence probability are highly sensitive to this setting. In contrast, maxlike directly estimates prevalence."

```{r carw.data, echo=FALSE}
# Carolina Wren data used in Royle et. al (2012) 
data(carw)
# Covert data.frame to a list of rasters 
rl <- lapply(carw.data$raster.data, function(x) { 
  m <- matrix(x, nrow=carw.data$dim[1], ncol=carw.data$dim[2], byrow=TRUE) 
  r <- raster(m) 
  extent(r) <- carw.data$ext 
  r 
  })
# Create a raster stack and add layer names 
rs <- stack(rl[[1]], rl[[2]], rl[[3]], rl[[4]], rl[[5]], rl[[6]]) 
names(rs) <- names(carw.data$raster.data)
plot(rs)
# Fit a model 
fm <- maxlike(~pcMix + I(pcMix^2) + pcDec + I(pcDec^2)+ pcCon 
              + I(pcCon^2) + pcGr + I(pcGr^2) + Lat + I(Lat^2) 
              + Lon + I(Lon^2), rs, carw.data$xy1, 
              method="BFGS", removeDuplicates=TRUE, savedata=TRUE)
summary(fm) 
confint(fm) 
AIC(fm) 
logLik(fm)

# Produce species distribution map (ie, expected probability of occurrence) 
psi.hat <- predict(fm) # Will warn if savedata=FALSE 
plot(psi.hat) 
points(carw.data$xy1, pch=16, cex=0.1)

# MAXENT sets "default prevalence" to an arbitrary value, 0.5. 
# We could do something similar by fixing the intercept at logit(0.5)=0. 
# However, it seems more appropriate to estimate this parameter.

fm.fix <- update(fm, fixed=c(0, rep(NA,length(coef(fm))-1)))

# Predict data.frame 
presenceData <- as.data.frame(extract(rs, carw.data$xy1)) 
presenceData <- presenceData[complete.cases(presenceData), ] 
presence.predictions <- predict(fm, newdata=presenceData) 
summary(presence.predictions)

# Calibrate with data.frames 
PresenceUniqueCells <- unique(cellFromXY(rs, xy=carw.data$xy1)) 
PresenceUnique <- xyFromCell(rs, PresenceUniqueCells) 
presenceData <- as.data.frame(extract(rs, PresenceUnique)) 
library(dismo) 
background <- randomPoints(rs, n=ncell(rs), extf=1.00) 
backgroundData <- as.data.frame(extract(rs, y=background)) 
backgroundData <- backgroundData[complete.cases(backgroundData), ] 
fm2 <- maxlike(~pcMix + I(pcMix^2) + pcDec + I(pcDec^2)+ pcCon 
              + I(pcCon^2) + pcGr + I(pcGr^2) + Lat + I(Lat^2) 
              + Lon + I(Lon^2), 
              rasters=NULL, points=NULL, 
              x=presenceData, z=backgroundData, 
              method="BFGS", removeDuplicates=TRUE, savedata=TRUE)
summary(fm2)
fm2$rasters <- rs 
psi.hat2 <- predict(fm2)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
